---
title: "A random walk model of Persson et al data"
output:
  html_document:
    df_print: paged
---

This is a rough reanalysis of Persson et al (2016)'s analysis of the ICCS2009 data. Comments and citations will be added later.

```{r setup}
library(foreign)
library(dplyr)
library(tidyr)
library(lubridate)
library(cmdstanr)
library(HDInterval)
library(posterior)

library(ggplot2)
theme_set(theme_classic())

options(mc.cores = 4)
```

Functions for scaling the IGMRFs. Functions `rw1_scale()` and `rw2_scale()` based on https://mc-stan.org/users/documentation/case-studies/icar_stan.html

```{r}
rw2_Q <- function(n) {
  if(n < 4)
    stop("Random walk must contain 4 or more points.")
  
  m <- Matrix::Diagonal(n)
  Matrix::diag(m) <- c(1, 5, rep(6, n-4), 5, 1)
  m[row(m) - col(m) == 1] <- m[row(m) - col(m) == -1] <- c(-2, rep(-4, n-3), -2)
  m[abs(row(m) - col(m)) == 2] <- 1
  
  return(m)
}

rw2_scale <- function(n, pc = NULL) {
  Q <- rw2_Q(n)
  
  Q_pert <- Q + Matrix::Diagonal(n) * max(Matrix::diag(Q)) * sqrt(.Machine$double.eps)  # usual perturbation

  if(is.null(pc)) {
    Q_inv <- INLA::inla.qinv(
      Q_pert,
      constr = list(A = matrix(c(1:n, rep(1, n)), nrow = 2, byrow = TRUE), e = c(0, 0))
    )
  } else {
    temp <- rep(0, n)
    temp[pc] <- 1
    Q_inv <- INLA::inla.qinv(
      Q_pert,
      constr = list(A = matrix(c(1:n, rep(1, n), temp), nrow = 3, byrow = TRUE), e = c(0, 0, 0))
    )
  }

  Q_diag <- Matrix::diag(Q_inv)
  Q_diag[pc] <- 0
  
  if(any(Q_diag < 0))
    stop("Covariance matrix not positive definite.")
  
  Q_diag <- Q_diag[Q_diag > 0]
  exp(mean(log(Q_diag)))  # geometric mean of the variances ("generalized variance")
}

rw1_Q <- function(n) {
  if(n < 2)
    stop("Random walk must contain 2 or more points.")
  
  m <- Matrix::Diagonal(n)
  Matrix::diag(m) <- c(1, rep(2, n-2), 1)
  m[abs(row(m) - col(m)) == 1] <- -1
  
  return(m)
}

rw1_scale <- function(n, pc = NULL) {
  Q <- rw1_Q(n)
  
  Q_pert <- Q + Matrix::Diagonal(n) * max(Matrix::diag(Q)) * sqrt(.Machine$double.eps)  # usual perturbation'
  
  if(is.null(pc)) {
    Q_inv <- INLA::inla.qinv(Q_pert, constr = list(A = matrix(rep(1, n), nrow = 1), e = 0))
  } else {
    temp <- rep(0, n)
    temp[pc] <- 1
    Q_inv <- INLA::inla.qinv(
      Q_pert,
      constr = list(A = matrix(c(rep(1, n), temp), nrow = 2, byrow = TRUE), e = c(0, 0))
    )
  }
  
  Q_diag <- Matrix::diag(Q_inv)
  Q_diag[pc] <- 0
  
  if(any(Q_diag < 0))
    stop("Covariance matrix not positive definite.")
  
  Q_diag <- Q_diag[Q_diag > 0]
  exp(mean(log(Q_diag)))  # geometric mean of the variances ("generalized variance")
}

rw1_scale(24)
rw1_scale(24, 13)
rw2_scale(24)
rw2_scale(24, 13)
```

# Importing ICCS2009 data

We process the data as in Persson et al (2016). I wrote this a year ago and should probably rewrite it more concisely.

```{r}
grc8 <- read.spss(
  "D:\\Data\\ICCS2009_IDB_SPSS\\Data_G8\\ISGGRCC2.sav",
  add.undeclared.levels = "no",
  to.data.frame = TRUE
)
nor8 <- read.spss(
  "D:\\Data\\ICCS2009_IDB_SPSS\\Data_G8\\ISGNORC2.sav",
  add.undeclared.levels = "no",
  to.data.frame = TRUE
)
svn8 <- read.spss(
  "D:\\Data\\ICCS2009_IDB_SPSS\\Data_G8\\ISGSVNC2.sav",
  add.undeclared.levels = "no",
  to.data.frame = TRUE
)
grc9 <- read.spss(
  "D:\\Data\\ICCS2009_IDB_SPSS\\Data_G9\\JSGGRCC2.sav",
  add.undeclared.levels = "no",
  to.data.frame = TRUE
)
nor9 <- read.spss(
  "D:\\Data\\ICCS2009_IDB_SPSS\\Data_G9\\JSGNORC2.sav",
  add.undeclared.levels = "no",
  to.data.frame = TRUE
)
svn9 <- read.spss(
  "D:\\Data\\ICCS2009_IDB_SPSS\\Data_G9\\JSGSVNC2.sav",
  add.undeclared.levels = "no",
  to.data.frame = TRUE
)
iccs8 <- rbind(grc8, nor8, svn8)
iccs9 <- rbind(grc9, nor9, svn9)
df8 <- data.frame(
  nwlciv = iccs8$NWLCIV,
  birth_month = -round(12 * (iccs8$ITDATEY + iccs8$ITDATEM/12 - iccs8$SAGE - 1995)),
  country = c(
    rep("Greece", nrow(grc8)),
    rep("Norway", nrow(nor8)),
    rep("Slovenia", nrow(svn8))
  )
)
df8$grade9 <- rep(0, nrow(df8))
df9 <- data.frame(
  nwlciv = iccs9$NWLCIV,
  birth_month = -round(12 * (iccs9$ITDATEY + iccs9$ITDATEM/12 - iccs9$SAGE - 1995)),
  country = c(
    rep("Greece", nrow(grc9)),
    rep("Norway", nrow(nor9)),
    rep("Slovenia", nrow(svn9))
  )
)
df9$grade9 <- rep(1, nrow(df9))

# Sweden
iccs8S <- read.spss(
  "D:\\Data\\ICCS2009_IDB_SPSS\\Data_G8\\ISGSWEC2.sav",
  add.undeclared.levels = "no",
  to.data.frame = TRUE
)
iccs9S <- read.spss(
  "D:\\Data\\ICCS2009_IDB_SPSS\\Data_G9\\JSGSWEC2.sav",
  add.undeclared.levels = "no",
  to.data.frame = TRUE
)
df8S <- data.frame(
  nwlciv = iccs8S$NWLCIV,
  birth_month = -round(12 * (iccs8S$ITDATEY + iccs8S$ITDATEM/12 - iccs8S$SAGE - 1994)),
  country = rep("Sweden", nrow(iccs8S))
)
df8S$grade9 <- rep(0, nrow(df8S))
df9S <- data.frame(
  nwlciv = iccs9S$NWLCIV,
  birth_month = -round(12 * (iccs9S$ITDATEY + iccs9S$ITDATEM/12 - iccs9S$SAGE - 1994)),
  country = rep("Sweden", nrow(iccs9S))
)
df9S$grade9 <- rep(1, nrow(df9S))

# combine data
df <- 
  rbind( df8, df9, df8S, df9S ) %>% 
  filter(
    !is.na(country),
    !is.na(nwlciv),
    birth_month <= 11,
    birth_month >= -12
  ) %>% 
  mutate( cutoff = ifelse(birth_month < 0, 0, 1) ) %>% 
  filter( grade9 == cutoff ) %>% 
  select( -cutoff )
attr(df$nwlciv, "value.labels") <- NULL
attr(df$birth_month, "value.labels") <- NULL
str(df)
```

Pick a single country.

```{r}
df_country <- 
  df %>% 
  filter(country == "Sweden") %>% 
  mutate(month_id = birth_month - min(birth_month) + 1) %>% 
  mutate(across(c(grade9, month_id), as.integer))

df_country
```

```{r}
df_country %>% 
  ggplot() +
  geom_point(aes(x = birth_month, y = nwlciv), alpha = 0.1)
```

Prep the data for the aggregate-Gaussian likelihood.

```{r}
df_agg <- 
  df_country %>% 
  group_by(birth_month, month_id, grade9) %>% 
  summarise(
    nwlciv_m = mean(nwlciv),
    nwlciv_v = var(nwlciv),
    students = n(),
    .groups = "drop"
  )

df_agg
```

```{r}
rw_mod <- cmdstan_model("rw2iid_pc_aggnormal.stan")
```

```{r}
scaling_factor <- rw2_scale(max(df_agg$month_id), 13)

fit <- rw_mod$sample(
  data = list(
    N_data = nrow(df_agg),
    x = df_agg$birth_month,
    y_m = df_agg$nwlciv_m,
    y_v = df_agg$nwlciv_v,
    y_n = df_agg$students,
    N_x = max(df_agg$month_id),
    x_id = df_agg$month_id,
    grade9 = df_agg$grade9,
    igmrf_order = 2,
    scaling_factor = scaling_factor,
    igmrf_pc = 13,
    prior_only = 0
  ),
  iter_warmup = 1e3,
  iter_sampling = 1e3,
  chains = 4,
  parallel_chains = 4,
  max_treedepth = 20
)
```

```{r}
fit$summary(c("Intercept", "b_g9", "b_x", "re_scale", "rho", "log_sigma")) %>% 
  mutate_if(is.numeric, round, 2)
```

```{r}
draws <- 
  fit$draws() %>% 
  as_draws_df() %>% 
  merge_chains()

plot(density(draws$re_scale))
```

```{r}
plot(density(draws$b_g9))
```

```{r}
plot(density(draws$rho, adj = 0.1))
curve(dbeta(x, 0.5, 0.5), lty = 2, add = TRUE)
```

```{r}
month_seq <- min(df_agg$birth_month):max(df_agg$birth_month)

draws %>% 
  subset_draws("mu") %>% 
  select(-.chain, -.iteration) %>% 
  pivot_longer(-.draw, names_to = "month", values_to = "mu") %>% 
  mutate(
    month = month %>% stringr::str_remove_all("[^0-9]") %>% as.integer(),
    month = month_seq[month]
  ) %>% 
  filter(.draw %in% sample(1:max(.draw), 200)) %>% 
  ggplot() +
  geom_line(aes(month, mu, group = .draw), alpha = 0.2)
```

```{r}
mu <- 
  draws %>% 
  subset_draws("mu") %>% 
  select(-.chain, -.iteration, -.draw)

mu_smry <- tibble(
  mu_mean = apply(mu, 2, mean),
  mu_l = apply(mu, 2, hdi)[1,],
  mu_u = apply(mu, 2, hdi)[2,],
  birth_month = month_seq
)

mu_smry
```

```{r}
ggplot() +
  geom_point(data = mu_smry, aes(birth_month, mu_mean), size = 4, color = "blue", alpha = 0.7) +
  geom_segment(
    data = mu_smry,
    aes(x = birth_month, xend = birth_month, y = mu_l, yend = mu_u),
    size = 2,
    color = "blue",
    alpha = 0.7
  )
```

Just the IGMRF component:

```{r}
month_id_seq <- seq_along(month_seq)

igmrf <- 
  draws %>% 
  subset_draws("igmrf") %>% 
  select(-.chain, -.iteration, -.draw)

mu_i <- sapply(
  month_id_seq,
  function(m)
    draws$re_scale*(sqrt(draws$rho/scaling_factor)*pull(igmrf, m))
)

mu_smry_i <- tibble(
  mu_mean = apply(mu_i, 2, mean),
  mu_l = apply(mu_i, 2, hdi)[1,],
  mu_u = apply(mu_i, 2, hdi)[2,],
  birth_month = month_seq
)

ggplot() +
  geom_point(data = mu_smry_i, aes(birth_month, mu_mean), size = 4, color = "blue", alpha = 0.7) +
  geom_segment(
    data = mu_smry_i,
    aes(x = birth_month, xend = birth_month, y = mu_l, yend = mu_u),
    size = 2,
    color = "blue",
    alpha = 0.7
  )
```

Heterogeneous only:

```{r}
hetero <- 
  draws %>% 
  subset_draws("hetero") %>% 
  select(-.chain, -.iteration, -.draw)

mu_h <- sapply(
  month_id_seq,
  function(m)
    draws$re_scale*(sqrt(1 - draws$rho)*pull(hetero, m))
)

mu_smry_h <- tibble(
  mu_mean = apply(mu_h, 2, mean),
  mu_l = apply(mu_h, 2, hdi)[1,],
  mu_u = apply(mu_h, 2, hdi)[2,],
  birth_month = month_seq
)

ggplot() +
  geom_point(data = mu_smry_h, aes(birth_month, mu_mean), size = 4, color = "blue", alpha = 0.7) +
  geom_segment(
    data = mu_smry_h,
    aes(x = birth_month, xend = birth_month, y = mu_l, yend = mu_u),
    size = 2,
    color = "blue",
    alpha = 0.7
  )
```

Fit against sample means of the data:

```{r fig.width=9}
ggplot() +
  geom_point(data = mu_smry, aes(birth_month, mu_mean), size = 4, color = "blue", alpha = 0.7) +
  geom_segment(
    data = mu_smry,
    aes(x = birth_month, xend = birth_month, y = mu_l, yend = mu_u),
    size = 2,
    color = "blue",
    alpha = 0.7
  ) +
  geom_point(data = df_agg, aes(birth_month, nwlciv_m)) +
  geom_segment(
    data = df_agg,
    aes(
      x = birth_month,
      xend = birth_month,
      y = nwlciv_m - 2*sqrt(nwlciv_v)/sqrt(students),
      yend = nwlciv_m + 2*sqrt(nwlciv_v)/sqrt(students)
    )
  )
```


















































